When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-03-01 paper/paper.tex
Remove unnecessary commands

\newcommand{\citet}[1]{\citeA{#1}}
\newcommand{\citep}[1]{\cite{#1}}

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete \citep{DBLP:books/fm/GareyJ79}, modern subgraph
isomorphism algorithms based upon constraint programming techniques can handle problem instances
with many hundreds of vertices in the pattern graph, and up to ten thousand vertices in the target
graph
\citep{DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
and subgraph isomorphism is used successfully in application areas including computer vision
\citep{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\citep{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\citep{DBLP:journals/ijprai/ConteFSV04}.  A labelled version of subgraph isomorphism
is also one of the key components in supporting complex queries in graph databases, which we discuss in
detail in \cref{section:filterverify}.

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated pairs of pattern and target graphs, sometimes together with a mix of real-world graphs,
for example biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. It also allows
experimenters to control graph parameters such as order, density, or degree distribution, and to
study scaling properties of algorithms with respect to these parameters.  The random instances used
in previous evaluations come from common datasets
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh) graph
and permuting the vertices. Such instances are guaranteed to be satisfiable---\citet{o:AntonO09}
exploited this property to create large sets of random satisfiable boolean satisfiability instances.
This is the most common approach to generating random subgraph isomorphism instances, meaning
existing benchmark suites contain relatively few non-trivial unsatisfiable instances (although a few
of the patterns in the instances by \citeauthor{DBLP:journals/constraints/ZampelliDS10}\ have had
extra edges added, making them unsatisfiable). Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has led to bias in algorithm design, to the extent that some proposed techniques, such
as those of \citet{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

@@

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and
\citet{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism in Erd\H{o}s-R\'enyi random graphs, we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable, and
we see a solver-independent complexity peak near this phase transition. Additionally, understanding
this behaviour helps us to design better search strategies, by improving variable- and
value-ordering heuristics (that is, by improving the order in which pattern and target vertices are
selected for guessed assignments). This is the second contribution of this paper.

In \cref{section:induced} we look at the induced subgraph isomorphism problem.  For certain choices of
parameters, there are two phase transitions, going from satisfiable to unsatisfiable, and then from
unsatisfiable back to satisfiable. Again, when going from satisfiable to unsatisfiable (from either
direction), instances go from being trivial to really hard to solve. However, each of the three
solvers we test also finds the central unsatisfiable region to be hard, despite it not being near a
phase transition. To show that this is not a simple weakness of current subgraph isomorphism
algorithms, we verify that this region is also hard for boolean satisfiability, pseudo-boolean, and
mixed integer solvers, and under reduction to the clique problem. Interestingly, the constrainedness
measure proposed by \citet{DBLP:conf/aaai/GentMPW96} \emph{does} predict this difficult region---the
third contribution of this paper is to use these instances to provide evidence in favour of
constrainedness, rather than proximity to a phase transition, being an accurate predictor of
difficulty, and to show that constrainedness is not simply a refinement of a phase transition
prediction.

What about other random graph models? In \cref{section:regular} we see that graphs where every vertex
has the same degree (known as $k$-regular) exhibit very similar characteristics.  However, when
labels on vertices are introduced (as is commonly seen in real-world applications and in graph
database systems), richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and the popular VF2 algorithm \citep{DBLP:journals/pami/CordellaFSV04} behaves much worse
than two more recent constraint-based algorithms
\citep{DBLP:journals/ai/Solnon10,DBLP:conf/cp/McCreeshP15} in certain cases. In
\cref{section:labelled} we take a close look at these cases, and argue that they \emph{should} be
easy to solve.

@@

Parts of this paper were presented at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify} are new material. Most of this work
also appeared in a PhD thesis \citep{o:McCreesh17}.

@@

An \emph{Erd\H{o}s-R\'enyi random graph}, written $G(n, p)$, is a graph with $n$ vertices, and an
edge between each distinct unordered pair of vertices with independent probability $p$
\citep{o:Gilbert59,o:ErdosR59}.\footnote{Note that elsewhere, this same name sometimes instead
refers to a graph $G(n, E)$ chosen uniformly from among all graphs with $n$ vertices and exactly $E$
edges.}

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes
RAM, running Scientific Linux release 6.7. We will be working with four subgraph isomorphism
solvers: the Glasgow solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9 with the ``-O3'' option.
The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem.

@@

This model can be extended with additional constraints. These allow more inference to be performed,
and so a solution may be found (or the lack of solution may be proven) with less search.  The
simplest example is that a pattern vertex of degree $d$ may only be mapped to a target vertex of
degree at least $d$, and so constraints can be posted which eliminate values corresponding to target
vertices of low degree from domains corresponding to pattern vertices of high degree. In the example
in \cref{figure:cpsipexample}, this rule would eliminate the values $2$, $3$ and $6$ from every domain at
the top of search. Richer constraints involving neighbourhood degree sequences
\citep{DBLP:journals/constraints/ZampelliDS10}, neighbourhood difference constraints
\citep{DBLP:journals/ai/Solnon10}, and distances and paths
\citep{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16} are also
possible.

@@

The line plots arithmetic mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial; for dense patterns, proving unsatisfiability is not particularly difficult; and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure~1 of \citet{DBLP:journals/cacm/Leyton-BrownHHX14}. In
particular, satisfiable instances tend to be easier, but show greater variation than unsatisfiable
instances, and there are exceptionally hard satisfiable instances \citep{DBLP:conf/cp/SmithG97}.

@@

This prediction is generally reasonably accurate, except that for very low and very high pattern
densities, we overestimate the satisfiable region. This is due to variance: although an expected
number of solutions much below one implies a high likelihood of unsatisfiability, it is not true
that a high expected number of solutions implies that any particular instance is likely to be
satisfiable. (Consider, for example, a sparse graph which has several isolated vertices. If one
solution exists, other symmetric solutions can be obtained by permuting the isolated vertices.
Thus although the expected number of solutions may be one, there cannot be exactly one solution.) A
similar behaviour is seen with random constraint satisfaction problems
\citep{DBLP:journals/ai/SmithD96}.

@@

The Glasgow and LAD solvers are driven by ordering heuristics: variable-ordering tells us which
pattern vertex to branch on, whilst value-ordering tells the order in which we should try target
vertices for that pattern vertex.
Various general principles have been considered when designing variable- and value-ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{DBLP:conf/cp/GentMPSW96}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation (i.e.\ $t^{\underline{p}}$) by as little
as possible.  When two or more domains are of equal size, LAD breaks ties lexicographically, whereas
Glasgow will pick a variable corresponding to a pattern vertex of highest degree. This strategy was
determined empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking
a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value-ordering heuristic, but Glasgow does: it prefers target vertices of highest degree.
Again, this was determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by
leaving as many vertices as possible available for future use. The search strategy used by VF2, in
contrast, is based around preserving connectivity, which gives very little discrimination except on
the sparsest of inputs.

@@

For patterns of order 10, it is always better to try to move towards the satisfiable region: if we
are in the bottom right diagonal half, we are best retaining the original heuristics (which move us
towards the top left), and if we are in the top left we should use the complement instead. This
goes against a suggestion by \citet{DBLP:conf/aaai/Walsh98} that switching heuristics based upon an estimate of
the solubility of the problem may offer good performance.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target vertex
$w$, we have a binary variable which takes the value 1 if and only if $v$ is mapped to $w$.
Constraints are added to ensure that each pattern vertex maps to exactly one target vertex, that
each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are mapped to
adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We use the
Clasp solver \citep{DBLP:journals/aicom/GebserKKOSS11} version 3.1.3 to solve the pseudo-boolean instances.  The
instances that are hard for the Glasgow solver remain hard for the PB solver, including instances
inside the central region, and the easy satisfiable instances remain easy. Similar results are seen
with the Glucose SAT solver \citep{o:glucose} using a direct encoding.  We also show an integer
program encoding: the Gurobi solver is only able to solve some of the trivial satisfiable instances,
and was almost never able to prove unsatisfiability within the time limit.

The \emph{association graph encoding} of a subgraph isomorphism problem is a reduction to the clique
decision problem. \citet{DBLP:conf/cp/McCreeshNPS16} describe and study this approach in more
detail. Briefly, the association graph is constructed by creating a new graph with a vertex for each
pair $(p, t)$ of vertices from the pattern and target graphs respectively. There is an edge between
vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$ to $t_1$ and $p_2$ to $t_2$
simultaneously is permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and only if $t_1$ is adjacent to
$t_2$. A clique of size equal to the order of the pattern graph exists in the association graph if
and only if the problem is satisfiable \citep{o:Levi73}. We used this encoding with an
implementation of \citet{DBLP:journals/cor/SegundoRJ11}'s bit-parallel maximum clique algorithm
BBMC, modified to solve the decision problem rather than the optimisation problem.  Again, our
results show that the instances in the central region remain hard, and additionally, some of the
easy unsatisfiable instances become hard.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \citep{DBLP:conf/aaai/GentMPW96}. An ensemble of problem instances with $\kappa < 1$ is
said to be underconstrained, and is likely to be made up of satisfiable instances; an ensemble with
$\kappa > 1$ is overconstrained, and is likely to be made of unsatisfiable instances. Empirically,
problem instances from an ensemble with $\kappa$ close to 1 are hard, and problem instances drawn
from an ensemble where $\kappa$ is very small or very large are usually easy.

@@

By $R(n, k)$ we mean a random graph with $n$ vertices, each of which has degree $k$. In
\cref{figure:kr} we recreate \cref{figure:non-induced}, using the NetworkX implementation
\citep{o:NetworkX} of the \citet{DBLP:journals/cpc/StegerW99} algorithm to generate regular graphs.
The $x$-axes range from degree 0 to degree $n - 1$ where $n \in \{ 10, 20, 30 \}$ is the number of
pattern vertices, and the y-axes range from degree 0 to degree
149 (one less than the number of target vertices).  The satisfiable / unsatisfiable plots show very
similar results to the non-induced problem on Erd\H{o}s-R\'enyi random graphs, with a similarly
sharp phase transition, and again only instances near the phase transition are difficult.

@@

Without labels, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$
possible injective assignments of target vertices to pattern vertices.  With labels, observe that
for any label $x$, vertices in $\operatorname{V}(P)|_x$ may only be mapped to vertices in
$\operatorname{V}(T)|_x$.  Thus for each label $x$, we have an expected $\nicefrac{p}{k}$ variables,
each of whose domains contain $\nicefrac{t}{k}$ values. We would like to say that the size of the
state space is now \[ |S| = \left((\nicefrac{t}{k})^{\underline{\nicefrac{p}{k}}}\right)^{k}
\textnormal{,} \] but to do this we must state what $a^{\underline{b}}$ means when $b$ is
fractional. The gamma function $\Gamma(n)$ is equal to $(n - 1)!$ for integers $n \ge 1$
\citep{Davis59}, but is also defined for positive real numbers, obeying the identify $\Gamma(x + 1)
= x\Gamma(x)$. By noting that $t^{\underline{p}} = \nicefrac{t!}{(t - p)!}$, we may obtain a
reasonable continuous extension by taking \[ |S| = \left(\frac{\Gamma\left(\nicefrac{t}{k} +
1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} + 1\right)}\right)^{k} \textnormal{\,.} \]

@@

So how good are these predictions? The black lines on the first row of heatmaps in
\cref{figure:labels} plot where we calculate $\langle Sol \rangle = 1$ will occur. For small numbers
of labels, our predictions are slightly better than in the unlabelled case: there seems to be less
of a variance problem for very dense patterns. Even as the number of labels becomes relatively
large, the prediction of the phase transition still occurs in the right place, but with ten labels
we start to see sporadic unsatisfiable instances deep inside the satisfiable region. With twenty
labels, we instead get a kind of phase transition from ``all unsatisfiable'' to ``mixed satisfiable
and unsatisfiable''. We can understand this intuitively: with sufficiently many labels, we might
generate, say, a red vertex adjacent to a blue vertex in the pattern, but not in the target.  With
twenty labels, we even sometimes generate no vertices with a particular label in the target at all.
This in many ways resembles ``flaws'' generated by certain random constraint satisfaction problem
instance generators
\citep{DBLP:journals/constraints/AchlioptasMKSKK01,DBLP:journals/constraints/GentMPSW01}.

@@

These experiments further highlight that VF2 occasionally finds some instances which should be easy hard.
But does this cause problems in practice? The literature suggests that it does. For example,
\citet{o:Gromping14} uses VF2 in a package for the R statistics language, and states
\begin{displayquote}``There are (not so many) instances for which creation of a clear design is
    prohibitively slow in the current implementation that evaluates subgraph isomorphism with the
    VF2 algorithm \ldots Recent experiences with a few of these showed that the LAD algorithm was
    very fast in ruling out impossible matches, where VF2 took a long time.''\end{displayquote}
Similarly, \citet{o:Murray12} uses VF2 inside a compiler, and observes extremely variable
(and prohibitively high) compile times in some cases due to the expense of subgraph isomorphism
calls---given the heavily labelled nature of these graphs, we conjecture that any domain-based
algorithm would eliminate this cost.

@@

An early graph database system by \citet{DBLP:conf/pods/ShashaWG02} uses a filtering heuristic to
eliminate unsatisfiable instances based upon simple structural elements. It is not clear whether the
aim is to reduce I/O costs, or to reduce the number of queries which must be tested, and the
experiments do not answer whether the indexing is effective. However, the work was influenced by a commercial
graph database system, whose documentation \citep{o:Daylight} states that indexing is
used to minimise disk accesses.

>>>>

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete \cite{DBLP:books/fm/GareyJ79}, modern subgraph
isomorphism algorithms based upon constraint programming techniques can handle problem instances
with many hundreds of vertices in the pattern graph, and up to ten thousand vertices in the target
graph
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
and subgraph isomorphism is used successfully in application areas including computer vision
\cite{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\cite{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\cite{DBLP:journals/ijprai/ConteFSV04}.  A labelled version of subgraph isomorphism
is also one of the key components in supporting complex queries in graph databases, which we discuss in
detail in \cref{section:filterverify}.

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated pairs of pattern and target graphs, sometimes together with a mix of real-world graphs,
for example biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. It also allows
experimenters to control graph parameters such as order, density, or degree distribution, and to
study scaling properties of algorithms with respect to these parameters.  The random instances used
in previous evaluations come from common datasets
\cite{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh) graph
and permuting the vertices. Such instances are guaranteed to be satisfiable---\citeA{o:AntonO09}
exploited this property to create large sets of random satisfiable boolean satisfiability instances.
This is the most common approach to generating random subgraph isomorphism instances, meaning
existing benchmark suites contain relatively few non-trivial unsatisfiable instances (although a few
of the patterns in the instances by \citeauthor{DBLP:journals/constraints/ZampelliDS10}\ have had
extra edges added, making them unsatisfiable). Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has led to bias in algorithm design, to the extent that some proposed techniques, such
as those of \citeA{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

@@

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citeA{DBLP:conf/ijcai/CheesemanKT91} and
\citeA{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism in Erd\H{o}s-R\'enyi random graphs, we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable, and
we see a solver-independent complexity peak near this phase transition. Additionally, understanding
this behaviour helps us to design better search strategies, by improving variable- and
value-ordering heuristics (that is, by improving the order in which pattern and target vertices are
selected for guessed assignments). This is the second contribution of this paper.

In \cref{section:induced} we look at the induced subgraph isomorphism problem.  For certain choices of
parameters, there are two phase transitions, going from satisfiable to unsatisfiable, and then from
unsatisfiable back to satisfiable. Again, when going from satisfiable to unsatisfiable (from either
direction), instances go from being trivial to really hard to solve. However, each of the three
solvers we test also finds the central unsatisfiable region to be hard, despite it not being near a
phase transition. To show that this is not a simple weakness of current subgraph isomorphism
algorithms, we verify that this region is also hard for boolean satisfiability, pseudo-boolean, and
mixed integer solvers, and under reduction to the clique problem. Interestingly, the constrainedness
measure proposed by \citeA{DBLP:conf/aaai/GentMPW96} \emph{does} predict this difficult region---the
third contribution of this paper is to use these instances to provide evidence in favour of
constrainedness, rather than proximity to a phase transition, being an accurate predictor of
difficulty, and to show that constrainedness is not simply a refinement of a phase transition
prediction.

What about other random graph models? In \cref{section:regular} we see that graphs where every vertex
has the same degree (known as $k$-regular) exhibit very similar characteristics.  However, when
labels on vertices are introduced (as is commonly seen in real-world applications and in graph
database systems), richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and the popular VF2 algorithm \cite{DBLP:journals/pami/CordellaFSV04} behaves much worse
than two more recent constraint-based algorithms
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/cp/McCreeshP15} in certain cases. In
\cref{section:labelled} we take a close look at these cases, and argue that they \emph{should} be
easy to solve.

@@

Parts of this paper were presented at IJCAI \cite{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify} are new material. Most of this work
also appeared in a PhD thesis \cite{o:McCreesh17}.

@@

An \emph{Erd\H{o}s-R\'enyi random graph}, written $G(n, p)$, is a graph with $n$ vertices, and an
edge between each distinct unordered pair of vertices with independent probability $p$
\cite{o:Gilbert59,o:ErdosR59}.\footnote{Note that elsewhere, this same name sometimes instead
refers to a graph $G(n, E)$ chosen uniformly from among all graphs with $n$ vertices and exactly $E$
edges.}

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes
RAM, running Scientific Linux release 6.7. We will be working with four subgraph isomorphism
solvers: the Glasgow solver \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\cite{DBLP:journals/ai/Solnon10}; VF2 \cite{DBLP:journals/pami/CordellaFSV04}; and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9 with the ``-O3'' option.
The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem.

@@

This model can be extended with additional constraints. These allow more inference to be performed,
and so a solution may be found (or the lack of solution may be proven) with less search.  The
simplest example is that a pattern vertex of degree $d$ may only be mapped to a target vertex of
degree at least $d$, and so constraints can be posted which eliminate values corresponding to target
vertices of low degree from domains corresponding to pattern vertices of high degree. In the example
in \cref{figure:cpsipexample}, this rule would eliminate the values $2$, $3$ and $6$ from every domain at
the top of search. Richer constraints involving neighbourhood degree sequences
\cite{DBLP:journals/constraints/ZampelliDS10}, neighbourhood difference constraints
\cite{DBLP:journals/ai/Solnon10}, and distances and paths
\cite{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16} are also
possible.

@@

The line plots arithmetic mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial; for dense patterns, proving unsatisfiability is not particularly difficult; and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure~1 of \citeA{DBLP:journals/cacm/Leyton-BrownHHX14}. In
particular, satisfiable instances tend to be easier, but show greater variation than unsatisfiable
instances, and there are exceptionally hard satisfiable instances \cite{DBLP:conf/cp/SmithG97}.

@@

This prediction is generally reasonably accurate, except that for very low and very high pattern
densities, we overestimate the satisfiable region. This is due to variance: although an expected
number of solutions much below one implies a high likelihood of unsatisfiability, it is not true
that a high expected number of solutions implies that any particular instance is likely to be
satisfiable. (Consider, for example, a sparse graph which has several isolated vertices. If one
solution exists, other symmetric solutions can be obtained by permuting the isolated vertices.
Thus although the expected number of solutions may be one, there cannot be exactly one solution.) A
similar behaviour is seen with random constraint satisfaction problems
\cite{DBLP:journals/ai/SmithD96}.

@@

The Glasgow and LAD solvers are driven by ordering heuristics: variable-ordering tells us which
pattern vertex to branch on, whilst value-ordering tells the order in which we should try target
vertices for that pattern vertex.
Various general principles have been considered when designing variable- and value-ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \cite{DBLP:conf/cp/GentMPSW96}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation (i.e.\ $t^{\underline{p}}$) by as little
as possible.  When two or more domains are of equal size, LAD breaks ties lexicographically, whereas
Glasgow will pick a variable corresponding to a pattern vertex of highest degree. This strategy was
determined empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking
a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value-ordering heuristic, but Glasgow does: it prefers target vertices of highest degree.
Again, this was determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by
leaving as many vertices as possible available for future use. The search strategy used by VF2, in
contrast, is based around preserving connectivity, which gives very little discrimination except on
the sparsest of inputs.

@@

For patterns of order 10, it is always better to try to move towards the satisfiable region: if we
are in the bottom right diagonal half, we are best retaining the original heuristics (which move us
towards the top left), and if we are in the top left we should use the complement instead. This
goes against a suggestion by \citeA{DBLP:conf/aaai/Walsh98} that switching heuristics based upon an estimate of
the solubility of the problem may offer good performance.

@@

The pseudo-boolean (PB) encoding is as follows. For each pattern vertex $v$ and each target vertex
$w$, we have a binary variable which takes the value 1 if and only if $v$ is mapped to $w$.
Constraints are added to ensure that each pattern vertex maps to exactly one target vertex, that
each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are mapped to
adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We use the
Clasp solver \cite{DBLP:journals/aicom/GebserKKOSS11} version 3.1.3 to solve the pseudo-boolean instances.  The
instances that are hard for the Glasgow solver remain hard for the PB solver, including instances
inside the central region, and the easy satisfiable instances remain easy. Similar results are seen
with the Glucose SAT solver \cite{o:glucose} using a direct encoding.  We also show an integer
program encoding: the Gurobi solver is only able to solve some of the trivial satisfiable instances,
and was almost never able to prove unsatisfiability within the time limit.

The \emph{association graph encoding} of a subgraph isomorphism problem is a reduction to the clique
decision problem. \citeA{DBLP:conf/cp/McCreeshNPS16} describe and study this approach in more
detail. Briefly, the association graph is constructed by creating a new graph with a vertex for each
pair $(p, t)$ of vertices from the pattern and target graphs respectively. There is an edge between
vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$ to $t_1$ and $p_2$ to $t_2$
simultaneously is permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and only if $t_1$ is adjacent to
$t_2$. A clique of size equal to the order of the pattern graph exists in the association graph if
and only if the problem is satisfiable \cite{o:Levi73}. We used this encoding with an
implementation of \citeA{DBLP:journals/cor/SegundoRJ11}'s bit-parallel maximum clique algorithm
BBMC, modified to solve the decision problem rather than the optimisation problem.  Again, our
results show that the instances in the central region remain hard, and additionally, some of the
easy unsatisfiable instances become hard.

@@

Constrainedness, denoted $\kappa$, is an alternative measure of difficulty designed to refine the
phase transition concept, and to generalise hardness parameters across different combinatorial
problems \cite{DBLP:conf/aaai/GentMPW96}. An ensemble of problem instances with $\kappa < 1$ is
said to be underconstrained, and is likely to be made up of satisfiable instances; an ensemble with
$\kappa > 1$ is overconstrained, and is likely to be made of unsatisfiable instances. Empirically,
problem instances from an ensemble with $\kappa$ close to 1 are hard, and problem instances drawn
from an ensemble where $\kappa$ is very small or very large are usually easy.

@@

By $R(n, k)$ we mean a random graph with $n$ vertices, each of which has degree $k$. In
\cref{figure:kr} we recreate \cref{figure:non-induced}, using the NetworkX implementation
\cite{o:NetworkX} of the \citeA{DBLP:journals/cpc/StegerW99} algorithm to generate regular graphs.
The $x$-axes range from degree 0 to degree $n - 1$ where $n \in \{ 10, 20, 30 \}$ is the number of
pattern vertices, and the y-axes range from degree 0 to degree
149 (one less than the number of target vertices).  The satisfiable / unsatisfiable plots show very
similar results to the non-induced problem on Erd\H{o}s-R\'enyi random graphs, with a similarly
sharp phase transition, and again only instances near the phase transition are difficult.

@@

Without labels, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$
possible injective assignments of target vertices to pattern vertices.  With labels, observe that
for any label $x$, vertices in $\operatorname{V}(P)|_x$ may only be mapped to vertices in
$\operatorname{V}(T)|_x$.  Thus for each label $x$, we have an expected $\nicefrac{p}{k}$ variables,
each of whose domains contain $\nicefrac{t}{k}$ values. We would like to say that the size of the
state space is now \[ |S| = \left((\nicefrac{t}{k})^{\underline{\nicefrac{p}{k}}}\right)^{k}
\textnormal{,} \] but to do this we must state what $a^{\underline{b}}$ means when $b$ is
fractional. The gamma function $\Gamma(n)$ is equal to $(n - 1)!$ for integers $n \ge 1$
\cite{Davis59}, but is also defined for positive real numbers, obeying the identify $\Gamma(x + 1)
= x\Gamma(x)$. By noting that $t^{\underline{p}} = \nicefrac{t!}{(t - p)!}$, we may obtain a
reasonable continuous extension by taking \[ |S| = \left(\frac{\Gamma\left(\nicefrac{t}{k} +
1\right)}{\Gamma\left(\nicefrac{t}{k} - \nicefrac{p}{k} + 1\right)}\right)^{k} \textnormal{\,.} \]

@@

So how good are these predictions? The black lines on the first row of heatmaps in
\cref{figure:labels} plot where we calculate $\langle Sol \rangle = 1$ will occur. For small numbers
of labels, our predictions are slightly better than in the unlabelled case: there seems to be less
of a variance problem for very dense patterns. Even as the number of labels becomes relatively
large, the prediction of the phase transition still occurs in the right place, but with ten labels
we start to see sporadic unsatisfiable instances deep inside the satisfiable region. With twenty
labels, we instead get a kind of phase transition from ``all unsatisfiable'' to ``mixed satisfiable
and unsatisfiable''. We can understand this intuitively: with sufficiently many labels, we might
generate, say, a red vertex adjacent to a blue vertex in the pattern, but not in the target.  With
twenty labels, we even sometimes generate no vertices with a particular label in the target at all.
This in many ways resembles ``flaws'' generated by certain random constraint satisfaction problem
instance generators
\cite{DBLP:journals/constraints/AchlioptasMKSKK01,DBLP:journals/constraints/GentMPSW01}.

@@

These experiments further highlight that VF2 occasionally finds some instances which should be easy hard.
But does this cause problems in practice? The literature suggests that it does. For example,
\citeA{o:Gromping14} uses VF2 in a package for the R statistics language, and states
\begin{displayquote}``There are (not so many) instances for which creation of a clear design is
    prohibitively slow in the current implementation that evaluates subgraph isomorphism with the
    VF2 algorithm \ldots Recent experiences with a few of these showed that the LAD algorithm was
    very fast in ruling out impossible matches, where VF2 took a long time.''\end{displayquote}
Similarly, \citeA{o:Murray12} uses VF2 inside a compiler, and observes extremely variable
(and prohibitively high) compile times in some cases due to the expense of subgraph isomorphism
calls---given the heavily labelled nature of these graphs, we conjecture that any domain-based
algorithm would eliminate this cost.

@@

An early graph database system by \citeA{DBLP:conf/pods/ShashaWG02} uses a filtering heuristic to
eliminate unsatisfiable instances based upon simple structural elements. It is not clear whether the
aim is to reduce I/O costs, or to reduce the number of queries which must be tested, and the
experiments do not answer whether the indexing is effective. However, the work was influenced by a commercial
graph database system, whose documentation \cite{o:Daylight} states that indexing is
used to minimise disk accesses.
